Multi-output Bus Travel Time Prediction\\ with Convolutional LSTM Neural Networks

### 2017-11-21 paper/paper.tex
Finish paper for review.

\documentclass[preprint,11pt,5p,twocolumn]{elsarticle}
\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{lineno}

@@

\makeatletter
\providecommand{\doi}[1]{%
  \begingroup
    \let\bibinfo\@secondoftwo
    \urlstyle{rm}%
    \href{http://dx.doi.org/#1}{%
      doi:\discretionary{}{}{}%
      \nolinkurl{#1}%
    }%
  \endgroup
}
\makeatother

@@

\title{Multi-output Bus Travel Time Prediction\\
Using Convolutional LSTM Neural Network}

@@

\author{Niklas Christoffer Petersen}

\address{
  Department of Management Engineering \\
  Technical University of Denmark, 2800 Kongens Lyngby \\
  niklch@dtu.dk
}

\begin{abstract}
Accurate and reliable predictions for travel times in a public transport networks is essential for delivering an attractive service.
This paper presents a multi-output, multi-time-step, deep neural network for bus travel time prediction using Convolutional and Long short-term memory (LSTM) layers. The method is evaluated and compared to other popular approaches for link travel time prediction, and currently available services.
\end{abstract}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Bus Travel Time Prediction \sep Intelligent Transport Systems \sep Convolutional Neural Network (CNN) \sep Long short-term memory (LSTM)
\end{keyword}

\end{frontmatter}
\section{Introduction}
Public transport authorities has long found that GPS trajectory data from already deployed \emph{Automatic Vehicle Location}-systems (AVL) can be used in \emph{Intelligent Transport Systems} (ITS) \cite{Tcrp48}. Examples include real-time traffic information for passengers, e.g.\ departure boards, where studies has shown that reliable real-time information at bus stops has a statistical significant dampening effect on the perceived waiting time \cite{Fan2016}.

Besides real-time passenger information, robust arrival- and departure time predictions are necessary for operating more sophisticated ITS applications successfully, e.g.\ demand-adaptive transit systems, and \emph{connection insurance} between different public transport services. Arrival/departure time prediction is commonly approached as a specialization of travel time prediction, where the predicted travel time is simply accumulated downstream the route to yield the arrival/departure time predictions at each stop point of the rest of the current journey. Besides the link travel time, estimations of dwell time (i.e.\ when a bus is holding at a stop point) are also accumulated downstream.

Producing precise bus travel time predictions in areas with little external influence, e.g.\ rural areas, can be solved to a large extent with historical averaging or simple regression methods \cite{Williams2003,Altinkaya2013}. The problem becomes much more complex in urban areas where congestion, events, road-works, weather, etc.\ highly influences the traffic flow and passenger demand.

In this paper we presents a multi-output, multi-time-step deep neural network for bus travel time prediction using a combination of \emph{Convolutional} and \emph{Long short-term memory} (LSTM) \cite{Lstm1,Lstm2} layers. The goal of this work is to produce precise short-term predictions (e.g.\ $0-1.5$ hours) for link travel time, specifically for bus traffic in urban areas. The method is evaluated and compared to other popular approaches for link travel time prediction, and public available services, e.g.\ Google Traffic.

The paper is structured in the following manner: In the next section related work and literature is reviewed. \Cref{sec:convlstm} introduces Convolutional LSTM neural networks in general and in \Cref{sec:model} we present the proposed multi-output model in more details, including e.g.\ network topology, etc. \Cref{sec:experiments} introduces the Copenhagen-dataset, which the model has been evaluated on, and our results are presented and discussed in~\Cref{sec:results}. Finally we conclude on the work in~\Cref{sec:conclusion}.

\subsection{Literature review}
Early approaches presents historical averaging models \cite{Dailey1999,Sun2007}, and linear regression \cite{Patnaik2004}. Recent research presents this type of model only for comparison purpose, and the models are in all cases outperformed by the compared models \cite{Shalaby2004,Jeong2005}. Kalman-filters, which by its capabilities of maintaining state between predictions, has shown interest from several studies, either as an independent model \cite{Chen2001,Shalaby2004} or in combination with other models \cite{Zaki2013,Bai2015}.

Both the different regression style models and the Kalman-filter models still has limited options for capturing fluctuations in travel and dwell time in a metropolitan bus system, since they to a large extent still is averaging and smoothing their response. E.g., the Kalman-filter's state is only directly accessible for the leading time-step, and thus not capable of finding long-distance patterns spanning over several links and/or over several time-steps. This is substantiated by \cite{Lin2013} and \cite{Kumar2014} that finds artificial neural networks (ANN) outperforms Kalman-filter models.

The computational challenges of pure ANN approaches i.a.\ has sparked the interest for studying composite or hybrid models. \citet{Zaki2013} uses a two-stage approach by combining offline ANN-models, with an adaptable/online Kalman-filter to yield a dynamic model. The model is not actually learning in the long term, but is able to adapt to temporal variations in the current travel time on a journey.

Some recent research recognizes that several routes can benefit from each other's predictions if they share some partial route segment, e.g.\ \cite{Yu2011,Gal2014,Bai2015}, however none of these approaches consider cross temporal correlations between different route segments, and only uses a small window for correlation with upstream links (e.g.\ max. 3 links).

Finally \citet{YanjieDuan2016} proposes to use LSTM model for general highway travel time prediction, and do predict multiple steps ahead, but only for a single link at a time, i.e.\ cross link (spacial) correlations are lost.

\section{Convolutional LSTM neural networks}
\label{sec:convlstm}
A Long short-term memory (LSTM) neural network is a special type of Recurrent Neural Network~(RNN) which has proven robust for capturing long-distance dependencies \cite{Lstm1,Lstm2}. The important feature of a LSTM network is its capability to persist cell state, $c_t$, from previous observations across sequences of input (e.g.\ time), but also eliminate information which is considered unimportant. To allow this mechanism the persistence of information is controlled by three gates: \emph{input gate}, \emph{forget gate}, and \emph{output gate}. Each gate yields a state value at time $t$, respectively $i_t$, $f_t$, and $o_t$, along with the cell output, $h_t$, cf.~\cref{eq:lstm}, where $\circ$ denotes the entry-wise product.

\begin{equation}
\begin{aligned}
  i_t &= \sigma \left( W_\mathit{xi} x_t + W_\mathit{hi} h_{t-1} + W_{ci} \circ c_{t-1} + b_i \right) \\
  f_t &= \sigma \left( W_\mathit{xf} x_t + W_\mathit{hf} h_{t-1} + W_\mathit{cf} \circ c_{t-1} + b_f \right) \\
  c_t &= f_t \circ c_{t-1} + i_t \circ \mathrm{tanh} \left( W_\mathit{xc} x_t + W_\mathit{hc} h_{t-1} + b_c \right) \\
  o_t &= \sigma \left( W_\mathit{xo} x_t + W_\mathit{ho} h_{t-1} + W_\mathit{co} \circ c_{t} + b_o \right) \\
  h_t &= o_t \circ \mathrm{tanh} \left( c_t \right)
\end{aligned}
\label{eq:lstm}
\end{equation}
\vspace{.5em}

\Cref{fig:lstm-peepholes} illustrates the inner structure of a LSTM cell with peephole as proposed by \cite{LstmPeephole}. 
Is has especially grown popular for predicting time series using methods evolved from \cite{LstmTs}, where fixed-length windows of time-series are generated and feed into a LSTM network. Multiple LSTMs can be stacked such more complex patterns of sequential information (e.g.\ temporal patterns) can be learned.

@@

Convolutional Neural Networks~(CNN) on the other hand has been widely used for capturing spacial relationships, e.g.\ importance of neighboring pixels in an image. As opposed to fully connected layers, where each unit, $i$, in the layer, has a dedicated scalar weight, $w_{ij}$, for all input value, $x_j$, convolutional units are only locally connected and reuses the same weights for producing several outputs. Instead of considering the entire input-vector, only a fixed sized window, or \emph{convolution}, around each input is considered. The weights are therefore referred to as the \emph{filters} or \emph{kernels} of the layer. \Cref{fig:conv} illustrates a single convolutional filter of size $3$ being applied to 1-dimensional data.

@@

Special care needs to be taken on the boundaries, i.e.\ where the convolutional filter will exceed the input. A popular approach to avoid the size of the output decreases is to pad the input, e.g.\ with zeros. This ensures that the output shape of each convolutional unit will always be identical to the input shape, which is often desirable. One of the key benefits of convolutional networks is that the number of weights that needs to be learned is considerably reduced compared to fully connected networks, and that learned patterns can be transfered across space. I.e.\ the convolutional filters become feature detectors, that in our case can detect spacial patterns across links, e.g.\ congestion forming, etc.

\citet{ConvLSTM} introduced the novel combination of Convolutional and LSTM layers into a single structure, the \emph{Convolutional LSTM}, or simply \emph{ConvLSTM}. Specifically the method applies convolutional filters in the \emph{input-to-state} and \emph{state-to-state} transitions the LSTM cf.~\cref{eq:convlstm}, where $*$ denotes the convolution operator.

\begin{equation}
\begin{aligned}
  i_t &= \sigma \left( W_\mathit{xi} * x_t + W_\mathit{hi} * h_{t-1} + W_{ci} \circ c_{t-1} + b_i \right) \\
  f_t &= \sigma \left( W_\mathit{xf} * x_t + W_\mathit{hf} * h_{t-1} + W_\mathit{cf} \circ c_{t-1} + b_f \right) \\
  c_t &= f_t \circ c_{t-1} + i_t \circ \mathrm{tanh} \left( W_\mathit{xc} * x_t + W_\mathit{hc} * h_{t-1} + b_c \right) \\
  o_t &= \sigma \left( W_\mathit{xo} * x_t + W_\mathit{ho} * h_{t-1} + W_\mathit{co} \circ c_{t} + b_o \right) \\
  h_t &= o_t \circ \mathrm{tanh} \left( c_t \right)
\end{aligned}
\label{eq:convlstm}
\end{equation}
\vspace{.5em}

The output dimensionality of a \emph{ConvLSTM} layer is like the traditional CNN-layer determined by the number of filters applied. However \emph{ConvLSTM} requires a total of eight filters for each desired output, i.e.\ four \emph{input-to-state} filters ($W_\mathit{xi}$, $W_\mathit{xf}$, $W_\mathit{xc}$, and $W_\mathit{xo}$) and four \emph{state-to-state} filters ($W_\mathit{hi}$, $W_\mathit{hf}$, $W_\mathit{hc}$, and $W_\mathit{ho}$).

% Hmm, i really would like this to get numbered before fig:data_shape, but cant make it work.
\afterpage{
\begin{figure*}[!t]
  \centering
  \includegraphics[scale=.75]{drawings/conv_lstm.pdf}
  \caption{Convolutional LSTM network topology.}
  \label{fig:ConvLSTM}
\end{figure*}
}
%\newpage

@@

\Cref{fig:ConvLSTM} shows the overall network topology, where blue boxes illustrates \emph{input-to-state} convolutions and yellow \emph{state-to-state} convolutions. The network uses an encoder/decoder technique that to a large extent follows \cite{ConvLSTM}, where the encoder-block consists of two \emph{ConvLSTM} layers. The result is feed into a decoder, or prediction block, also consisting of two \emph{ConvLSTM} layers. The architecture allows unequal $w$ and $k$, e.g.\ predict the next $3$ time-steps based on a window size of $20$.

@@

To avoid over-fitting doing training \emph{Dropout} \cite{Dropout} is performed between the \emph{ConvLSTM} layers, and \emph{Batch Normalization} \cite{BatchNorm} are also performed before each \emph{ConvLSTM} layer to ensure reasonable input for activations. The dropout factor is adjusted to respectively 20\%, 10\% and 10\%.

Each of the \emph{ConvLSTM} layers uses linear activation functions, and the output from the last layer in the decoder-block is feed into a fully connected (FC) layer using the \emph{ReLU} activation function, which also ensures only positive travel times are predicted.

@@

Likewise is the output arranged with $N$ predictions of $k$ time steps ahead, $t+1, \ldots, t+k$. Thus the input is a 4D-tensor, $X$ with shape $(N, w, u, 1)$, and the output, $Y$, a 4D-tensor with shape $(N, k, u, 1)$ -- in both bases the last one refers to the single link travel time for each time-step/link combination. It is emphasized that each prediction consists of travel time predictions for all links for the next $k$ time-steps, i.e.\ multi-output, multi-time-step prediction.

The $N$ samples are sampled at a fixed time resolution, since we need a shared time reference across all links. \Cref{sec:experiments} elaborates on some of the considerations in choosing a sound resolution.

\begin{figure}[!ht]
  \centering
  \includegraphics[scale=1.1]{drawings/data_shape.pdf}
  \caption{Shapes of the input and output data.}
  \label{fig:data_shape}
\end{figure}

\subsection{Detrending}
Urban bus travel times varies throughout the time of the day, and the day of the week due to \emph{recurring congestion}. In order to reduce the need for the network to learn this recurring variation, link travel for link $\mathit{ln} \in \{ 1,\ldots,u\}$, at time-step $t$, $x_{\mathit{ln},t}$, is normalized to focus on deviations from the normal and expected pattern. Travel times are centered with the mean for each link, at the time of day, and day of week, $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$, and scaled with the standard deviation for each link, $\sigma_\mathit{ln}$, cf.~\cref{eq:normalization}.

@@

When calculating the mean and standard deviation it can be beneficial to exclude extreme outliers, since both mean and standard deviation are higly sensitive to such measurements. A suggested method is to apply \emph{absolute deviation around the median} (MAD) cf.~\cite{Olewuezi2011} when calculating $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$ and $\sigma_\mathit{ln}$.

@@

During training the variables $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$ and $\sigma_\mathit{ln}$ should be calculated solely based on the training set, to emulated real-world application.

    \caption{Examples of travel time for a single link over a single day, at various time resolutions.}
    \label{fig:resolutions}
\end{figure*}
\newpage

@@

For evaluation the method is applied to a dataset from Copenhagen's public transport authority, ``Movia''. The dataset consists of 1,2~M~travel time observations for the ``4A'' bus line in the period May to October 2017. The data points were collected using the real-time AVL-system installed in every vehicle servicing the line.

@@

\subsection{Time resolution}
To allow predictions for fixed time-steps ahead, data is aggregated at a fixed time resolution. \Cref{fig:resolutions} shows examples of travel time for a single link over a single day at various time resolutions. The black dots are actual measurements, and the lines the aggregated mean link travel time at the given resolution. The choice depends of the \emph{expected} frequency of the line, and is a balance between capturing the details and still having a reasonal number of measurements in each time-step to avoid over-fitting.

For this experiment data was aggregated into 15-minute time-steps and normalized cf.~\Cref{sec:model}. This resolution were chosen based on a measured mean \emph{headway} of $7.5$ minutes between 06:00 and 22:00, i.e.\ the time between two vehicles during daytime.

%TODO: How much used for training, how much for testing?
\subsection{Evaluation}
In order to evaluate the presented model and comparisons, the following measures is used: \emph{mean absolute error} (MAE), \emph{root mean square error} (RMSE), and \emph{mean absolute percentage error} (MAPE) cf.~\cref{eq:mae,eq:rmse,eq:mape}, where $Y_i$ is the true link travel times for sample $i$ and $\widehat{Y}_i$ is the predicted travel time.

\begin{equation}
    \textrm{MAE}(Y, \widehat{Y}) = \frac{\sum_{i = 1}^{N} \left| Y_i - \widehat{Y}_i \right| }{N}
    \label{eq:mae}
\end{equation}

\begin{equation}
    \textrm{RMSE}(Y, \widehat{Y}) = \sqrt{\frac{\sum_{i = 1}^{N} \left(Y_i - \widehat{Y}_i \right)^2}{N}}
    \label{eq:rmse}
\end{equation}

\begin{equation}
    \textrm{MAPE}(Y, \widehat{Y}) = \frac{1}{N} \sum_{i = 1}^{N} \left| \frac{Y_i - \widehat{Y}_i}{Y_i} \right| 
    \label{eq:mape} 
\end{equation}
\vspace{.5em}

>>>>

\documentclass[preprint,11pt,5p,twocolumn]{elsarticle}
\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{xfrac}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{lineno}

@@

\newcommand{\matr}[1]{\mathbf{#1}}

\makeatletter
\providecommand{\doi}[1]{%
  \begingroup
    \let\bibinfo\@secondoftwo
    \urlstyle{rm}%
    \href{http://dx.doi.org/#1}{%
      doi:\discretionary{}{}{}%
      \nolinkurl{#1}%
    }%
  \endgroup
}

\let\oldhref\href
\renewcommand{\href}[2]{\oldhref{#1}{\hbox{#2}}}
\makeatother

@@

\title{Multi-output Bus Travel Time Prediction\\
with Convolutional LSTM Neural Networks}

@@

\author{Niklas Christoffer Petersen\\[.25em]\footnotesize{niklch@dtu.dk}}

\address{
  \vspace{.5em}
  \begin{minipage}{.5\textwidth}
    \center
    Department of Management Engineering \\
    Technical University of Denmark, 2800 Kongens Lyngby 
  \end{minipage}% This must go next to `\end{minipage}`
  \begin{minipage}{.5\textwidth}
    \center
    Public Transport Movia \\
    Gammel Køge Landevej 3, 2500 Valby 
  \end{minipage}
  \vspace{-2.5em}
}

\begin{abstract}
Accurate and reliable predictions for travel times in public transport networks is essential for delivering an attractive service.
This paper presents a multi-output, multi-time-step, deep neural network for bus travel time prediction using Convolutional and Long short-term memory (LSTM) layers. The method is empirically evaluated and compared to other popular approaches for link travel time prediction, and currently available services. We find that the proposed model significantly outperforms all the other methods, and is able to detect small irregular peeks in bus travel time very quickly.
\end{abstract}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Bus Travel Time Prediction \sep Intelligent Transport Systems \sep Convolutional Neural Network (CNN) \sep Long short-term memory (LSTM), Deep Learning.
\end{keyword}

\end{frontmatter}
\section{Introduction}
Public transport authorities have long found that GPS trajectory data from already deployed \emph{Automatic Vehicle Location}-systems (AVL) can be used in \emph{Intelligent Transport Systems} (ITS) \cite{Tcrp48}. Examples include real-time traffic information for passengers, e.g.\ smartphone travel planner apps, in-vehicle infotainment screens and departure boards. Studies has shown that reliable real-time information at bus stops has a statistical significant dampening effect on the perceived waiting time \cite{Fan2016}.

Besides real-time passenger information, robust arrival- and departure time predictions are necessary for operating more sophisticated ITS applications successfully, e.g. \emph{connection assurance} between low frequent public transport services, where accurate arrival times are needed to decide if one of the services should wait on the other. Another example is \emph{demand-adaptive transit systems}, where prolonged travel time should be taken into account because of the influence on both vehicle and driver utilization, and thus the efficiency of the transit system. Arrival/departure time prediction is commonly approached as a specialization of travel time prediction as illustrated in \Cref{fig:bus-arrival-links}. The predicted travel time for each link is simply accumulated downstream the route to yield the arrival/departure time predictions at each stop point of the rest of the current journey. Besides the link travel time, estimations of dwell time (i.e.\ when a bus is holding at a stop point) are also accumulated downstream.

\begin{figure}[!ht]
  \center  
  \includegraphics[width=\columnwidth]{drawings/bus-arrival-links-narrow.pdf}
  \caption{Arrival- and departure using link travel time.}
  \label{fig:bus-arrival-links}
\end{figure}

Producing precise bus travel time predictions in areas with little external influence, e.g.\ rural areas, can be solved to a large extent with historical averaging or simple regression methods \cite{Williams2003,Altinkaya2013}. The problem becomes much more complex in urban areas where congestion, events, road-works, weather, etc.\ highly influence the traffic flow and passenger demand. As on-board GPS and AVL-systems has become more affordable and common, data has both grown in coverage, i.e.\ number of vehicles with AVL installed, and frequency, i.e.\ number of GPS-positions collected per vehicle.

In this paper we present a multi-output, multi-time-step deep neural network for bus travel time prediction using a combination of \emph{Convolutional} and \emph{Long short-term memory} (LSTM) \cite{Lstm1,Lstm2} layers. The goal of this work is to produce precise short-term predictions (e.g.\ $0-1.5$ hours) for link travel time, specifically for bus traffic in urban areas. The method is empirically evaluated and compared to other popular approaches for link travel time prediction, including the currently model deployed in production by the public transport authority for the Greater Copenhagen Area: \emph{Movia}. Furthermore the method is compared to Google Traffic, a popular online service for travel time prediction.

This paper is structured in the following manner: In the next section, related work and literature are reviewed. \Cref{sec:convlstm} introduces Convolutional LSTM neural networks in general and in \Cref{sec:model} we present the proposed multi-output model in more details, including e.g.\ network topology, etc. \Cref{sec:experiments} introduces the Copenhagen-dataset, which the model has been evaluated on, and our results are presented and discussed in~\Cref{sec:results}. Finally we conclude on the work in~\Cref{sec:conclusion}.

\section{Literature review}
Bus link travel time prediction has been explored in research as GPS and AVL data has become increasingly available. The problem overlaps with other research areas such as general traffic flow and speed estimation, but also has unique constraints and opportunities that follows from servicing a fixed route. The improvement in computational power in the recent decades has gradually allowed more complex link travel time models with increased precision.

Early approaches for bus travel time prediction rely on historical averaging models \cite{Dailey1999,Sun2007}, and linear regression \cite{Patnaik2004}. Recent research presents this type of model only for comparison purposes, and the models are in all cases outperformed by the proposed models \cite{Shalaby2004,Jeong2005}. A historical averaging model will only slowly converge to changes in the travel time, which of cause is undesired with short, but highly impacting, external influences (e.g.\ a traffic incident or large event).

Kalman-filters, which by its capabilities of maintaining state between predictions, has shown interest from several studies for capturing \emph{changes} in the travel time, either as an independent model \cite{Chen2001,Shalaby2004} or in combination with other models \cite{Yo2010,Bai2015}. The Kalman-filter models still have limited options for capturing and forecasting the complex dynamics of travel and dwell time in a metropolitan bus system. E.g., the Kalman-filter's state is only directly accessible for the leading time-step, and thus not capable of finding long-distance patterns spanning over several links and/or over several time-steps. This is substantiated by \cite{Lin2013} and \cite{Kumar2014}, who find artificial neural networks (ANN) to outperform Kalman-filter models. However, the computational challenges of fully connected ANNs is also limiting the depth of the network and thus how complex patterns it can learn to recognize. This has sparked the interest for studying composite or hybrid models. \citet{Bai2015} uses a two-stage approach by combining offline ANN-models, with an adaptable/online Kalman-filter to yield a dynamic model. The model is not actually learning in the long term, but is able to adapt to temporal variations in the current travel time on a journey, and is still not able to recognize long distance patterns. The proposed model in this work uses \emph{Long short-term memory} cells (LSTM), and differs from existing research in bus link travel time prediction by combining the capability for persisting state-space over multiple time-steps, while allowing the deep neural network to be feasibly trainable.

Some recent research recognizes that several routes can benefit from each other's predictions if they share some partial route segment, e.g.\ \cite{Yu2011,Gal2014,Bai2015}, however none of these approaches consider cross temporal correlations between different route segments, and only use a small window for correlation with upstream links (e.g.\ max.\ 3 links). Likewise, \citet{YanjieDuan2016} proposes to use a LSTM model for general highway travel time prediction, and do predict multiple time-steps ahead, but only for a single link at a time, i.e.\ cross link (spatial) correlations are lost. In contrast the combination of both \emph{LSTM} cells and \emph{Convolutional} filters for bus travel time prediction, proposed in this paper, allows the learned patterns to generalize beyond a single link and time, i.e.\ multi-output and multi-time-step. Furthermore this reduces the computational complexity by magnitudes compared to fully connected ANNs capable of capturing similar complex patterns.

\section{Convolutional LSTM neural networks}
\label{sec:convlstm}
A Long short-term memory (LSTM) neural network is a special type of Recurrent Neural Network~(RNN) which has proven robust for capturing long-distance dependencies \cite{Lstm1,Lstm2}. The important feature of a LSTM network is its capability to persist cell state, $\matr{c}_t$, from previous observations across sequences of input (e.g.\ time), but also eliminate information which is considered irrelevant. To allow this mechanism the persistence of information is controlled by three gates: \emph{input gate}, \emph{forget gate}, and \emph{output gate}. Each gate yields a state value at time $t$, respectively $\matr{i}_t$, $\matr{f}_t$, and $\matr{o}_t$, along with the cell output, $\matr{h}_t$, cf.~\cref{eq:lstm}, where $\circ$ denotes the entry-wise product.

\begin{equation}
\begin{aligned}
  \matr{i}_t &= \sigma \left( \matr{W}^i \matr{x}_t + \matr{R}^i \matr{h}_{t-1} + \matr{U}^i \circ \matr{c}_{t-1} + \matr{b}^i \right) \\
  \matr{f}_t &= \sigma \left( \matr{W}^f \matr{x}_t + \matr{R}^f \matr{h}_{t-1} + \matr{U}^f \circ \matr{c}_{t-1} + \matr{b}^f \right) \\
  \matr{c}_t &= \matr{f}_t \circ \matr{c}_{t-1} + \matr{i}_t \circ \mathrm{tanh} \left( \matr{W}^c \matr{x}_t + \matr{R}^c \matr{h}_{t-1} + \matr{b}_c \right) \\
  \matr{o}_t &= \sigma \left( \matr{W}^\mathit{o} \matr{x}_t + \matr{R}^\mathit{o} \matr{h}_{t-1} + \matr{U}^\mathit{o} \circ \matr{c}_{t} + \matr{b}^o \right) \\
  \matr{h}_t &= \matr{o}_t \circ \mathrm{tanh} \left( \matr{c}_t \right)
\end{aligned}
\label{eq:lstm}
\end{equation}
\vspace{.5em}

\Cref{fig:lstm-peepholes} illustrates the inner structure of a LSTM cell with peephole as proposed by \cite{LstmPeephole}. 
Is has especially grown popular for predicting time series using methods evolved from \cite{LstmTs}, where fixed-length windows of time-series are generated and feed into a LSTM network. Multiple LSTMs can be stacked such that more complex patterns of sequential information (e.g.\ temporal patterns) can be learned.

@@

Convolutional Neural Networks~(CNN) on the other hand have been widely used for capturing spacial relationships, e.g.\ importance of neighboring pixels in an image. As opposed to fully connected layers, where each unit, $i$, in the layer, has a dedicated scalar weight, $w_{ij}$, for all input value, $x_j$, convolutional units are only locally connected and reuse the same weights for producing several outputs. Instead of considering the entire input-vector, only a fixed-size window, or \emph{convolution}, around each input is considered. The weights are therefore referred to as the \emph{filters} or \emph{kernels} of the layer. \Cref{fig:conv} illustrates a single convolutional filter of size $3$ being applied to 1-dimensional data.

@@

Special care needs to be taken on the boundaries, i.e.\ where the convolutional filter will exceed the input. A popular approach to avoid the size of the output decreases is to pad the input, e.g.\ with zeros. This ensures that the output shape of each convolutional unit will always be identical to the input shape, which is often desirable. One of the key benefits of convolutional networks is that the number of weights that needs to be learned is considerably reduced compared to fully connected networks, and that learned patterns can be transfered across space. I.e.\ the convolutional filters become feature detectors, that in our case can detect spatial patterns across links, e.g.\ congestion forming, etc.

\citet{ConvLSTM} introduced the novel combination of Convolutional and LSTM layers into a single structure, the \emph{Convolutional LSTM}, or simply \emph{ConvLSTM}. Specifically the method applies convolutional filters in the \emph{input-to-state} and \emph{state-to-state} transitions of the LSTM cf.~\cref{eq:convlstm}, where $*$ denotes the convolution operator.
\begin{equation}
\begin{aligned}
  \matr{i}_t &= \sigma \left( \matr{W}^i * \matr{x}_t + \matr{R}^i * \matr{h}_{t-1} + \matr{U}^i \circ \matr{c}_{t-1} + \matr{b}^i \right) \\
  \matr{f}_t &= \sigma \left( \matr{W}^f * \matr{x}_t + \matr{R}^f * \matr{h}_{t-1} + \matr{U}^f \circ \matr{c}_{t-1} + \matr{b}^f \right) \\
  \matr{c}_t &= \matr{f}_t \circ \matr{c}_{t-1} + \matr{i}_t \circ \mathrm{tanh} \left( \matr{W}^c * \matr{x}_t + \matr{R}^c * \matr{h}_{t-1} + \matr{b}_c \right) \\
  \matr{o}_t &= \sigma \left( \matr{W}^\mathit{o} * \matr{x}_t + \matr{R}^\mathit{o} * \matr{h}_{t-1} + \matr{U}^\mathit{o} \circ \matr{c}_{t} + \matr{b}^o \right) \\
  \matr{h}_t &= \matr{o}_t \circ \mathrm{tanh} \left( \matr{c}_t \right)
\end{aligned}
\label{eq:convlstm}
\end{equation}
\vspace{.5em}

\begin{figure*}[!t]
  \centering
  \includegraphics[scale=.75]{drawings/conv_lstm.pdf}
  \caption{Convolutional LSTM network topology.}
  \label{fig:ConvLSTM}
\end{figure*}

The output dimensionality of a \emph{ConvLSTM} layer is like the traditional CNN-layer determined by the number of filters applied. However, \emph{ConvLSTMs} require a total of eight filters for each desired output, i.e.\ four \emph{input-to-state} filters ($\matr{W}^i$, $\matr{W}^f$, $\matr{W}^c$, and $\matr{W}^o$) and four \emph{state-to-state} filters ($\matr{R}^i$, $\matr{R}^f$, $\matr{R}^c$, and $\matr{R}^o$).
Still it is important to emphasize that application of convolutional filters to the LSTM model greatly reduces the number of weights that needs to be learned, compared to a \emph{pure LSTM} approach. This allows for even deeper networks.
\newpage

@@

\Cref{fig:ConvLSTM} shows the overall network topology, where blue boxes illustrates \emph{input-to-state} convolutions and yellow \emph{state-to-state} convolutions. The network uses an encoder/decoder technique that to a large extent follows \cite{ConvLSTM}, where the encoder-block consists of two \emph{ConvLSTM} layers. The result is fed into a decoder, or prediction block, also consisting of two \emph{ConvLSTM} layers. The architecture allows unequal $w$ and $k$, e.g.\ predict the next $3$ time-steps based on a window size of $20$.

@@

To avoid over-fitting during training \emph{Dropout} \cite{Dropout} is performed between the \emph{ConvLSTM} layers, and \emph{Batch Normalization} \cite{BatchNorm} is also performed before each \emph{ConvLSTM} layer to ensure reasonable inputs for the activations. The dropout probability is adjusted to respectively 20\%, 10\% and 10\%.

Each of the \emph{ConvLSTM} layers uses linear activation functions, and the output from the last layer in the decoder-block is fed into a fully connected (FC) layer using the \emph{ReLU} activation function, which also ensures that only positive travel times are predicted.

@@

Likewise, the output is arranged with $N$ predictions of $k$ time steps ahead, $t+1, \ldots, t+k$. Thus the input is a 4D-tensor, $\matr{X}$ with dimensionality $N \times w \times u \times 1$, and the output, $\matr{Y}$, a 4D-tensor with dimensionality $N \times k \times u \times 1$ -- in both cases the last one refers to the single link travel time for each time-step/link combination. It is emphasized that each prediction consists of travel time predictions for all links for the next $k$ time-steps, i.e.\ multi-output, multi-time-step prediction.

The $N$ samples are sampled at a fixed time resolution, since we need a shared time reference across all links. \Cref{sec:experiments} elaborates on some of the considerations for choosing a sound resolution.

\begin{figure}[!ht]
  \centering
  \includegraphics[scale=1.0]{drawings/data_shape.pdf}
  \caption{Shapes of the input and output data.}
  \label{fig:data_shape}
\end{figure}

\subsection{Detrending}
Urban bus travel times vary throughout the time of the day, and the day of the week due to \emph{recurring congestion}. In order to reduce the need for the network to learn this recurring variation, link travel for link $\mathit{ln} \in \{ 1,\ldots,u\}$, at time-step $t$, $x_{\mathit{ln},t}$, is normalized to focus on deviations from the normal and expected pattern. Travel times are centered with the mean for each link, at the time of day, and day of week, $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$, and scaled with the standard deviation for each link, $\sigma_\mathit{ln}$, cf.~\cref{eq:normalization}.

@@

When calculating the mean and standard deviation it can be beneficial to exclude extreme outliers, since both mean and standard deviation are highly sensitive to such measurements. A suggested method is to apply \emph{absolute deviation around the median} (MAD) cf.~\cite{Olewuezi2011} when calculating $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$ and $\sigma_\mathit{ln}$.

@@

During training the variables $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$ and $\sigma_\mathit{ln}$ should be calculated solely based on the training set, to emulate the real-world application.

@@

For the purpose of evaluation, the proposed method is applied to a dataset from Copenhagen's public transport authority: \emph{Movia}. The dataset consists of 1,2~M~travel time observations for the ``4A'' bus line in the period May to October 2017. The data points were collected using the real-time AVL-system installed in every vehicle servicing the line.

@@

    \caption{Examples of travel time for a single link over a single day, at various time resolutions.}
    \label{fig:resolutions}
\end{figure*}

\subsection{Time resolution}
To allow predictions for fixed time-steps ahead, data is aggregated at a fixed time resolution. \Cref{fig:resolutions} shows examples of travel time for a single link over a single day at various time resolutions. The black dots are actual measurements, and the lines the aggregated mean link travel time at the given resolution. The choice depends of the \emph{expected} frequency of the line, and is a balance between capturing the details and still having a reasonable number of measurements of each time-step to avoid over-fitting.

For this experiment data was aggregated into 15-minute time-steps and normalized cf.~\Cref{sec:model}. This resolution was chosen based on a measured mean \emph{headway} of $7.5$ minutes between 06:00 and 22:00, i.e.\ the time between two vehicles during daytime.

Given the time resolution, we set the fixed windows size, $w = 32$, equivalent of 8 hours, to allow pattern in the morning peek affect patterns in the afternoon peek. We set $k = 3$ to allow predictions of up to 45 minutes into the future.

\subsection{Evaluation}
In order to evaluate the proposed model and comparisons, the following measures are used: \emph{mean absolute error} (MAE), \emph{root mean square error} (RMSE), and \emph{mean absolute percentage error} (MAPE) cf.~\cref{eq:mae,eq:rmse,eq:mape}, where $\matr{Y}_i$ is the true link travel times for sample $i$ and $\matr{\widehat{Y}}_i$ is the predicted travel times. Since the multi-output, multi-time-step model predicts link travel times for all $u$ links for the next $k$ time steps, $\matr{Y}_i$ and $\matr{\widehat{Y}}_i$ have both the dimensionality $w \times u \times 1$.

\begin{equation}
    \textrm{MAE}(\matr{Y}, \matr{\widehat{Y}}) = \frac{\sum_{i = 1}^{N} \left| \matr{Y}_i - \matr{\widehat{Y}}_i \right| }{N}
    \label{eq:mae}
\end{equation}

\begin{equation}
    \textrm{RMSE}(\matr{Y}, \matr{\widehat{Y}}) = \sqrt{\frac{\sum_{i = 1}^{N} \left(\matr{Y}_i - \matr{\widehat{Y}}_i \right)^2}{N}}
    \label{eq:rmse}
\end{equation}
